---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```


## Loading and preprocessing the data  
Unzip and load the data into memory

```{r, echo = TRUE}
setwd("C:/Users/keith_bailey/Documents/GitHub/RepData_PeerAssessment1")
unzip("activity.zip")
activity<-data.table::fread("activity.csv")
```

Cleanup the dates so they show date types in the dataframe  
```{r, echo = TRUE}
activity$date<-as.Date(lubridate::fast_strptime(activity$date, format="%Y-%m-%d"))

```


## What is mean total number of steps taken per day?
```{r, echo = TRUE}
#Load required packages
require(dplyr)
require(ggplot2)
require(lubridate)
```

```{r, echo = TRUE}
steps_per_day<-activity%>%
               filter(!is.na(steps))%>%
               group_by(date)%>%
               summarise(total_no_steps_per_day = sum(steps))

steps_per_day
```

###Plot daily step data in histogram and barplot to investigate

```{r, echo=TRUE}
#histogram - default number of bins
qplot(steps_per_day$total_no_steps_per_day, geom="histogram") + labs(x="Total Number of Steps per Day")
```

The number of bins need some manual work as default too many to be meaningful; I prefer to use a density plot to assess the data

```{r, echo = TRUE}
ggplot(steps_per_day, aes(x=total_no_steps_per_day)) + geom_density() + labs(x="Total Number of Steps per Day")

```

``` {r, echo = TRUE}
#mean steps taken each day
mean(steps_per_day$total_no_steps_per_day)

#median steps taken each day
median(steps_per_day$total_no_steps_per_day)
```

## What is the average daily activity pattern?

``` {r, echo = TRUE}
steps_per_interval<-activity%>%
               filter(!is.na(steps))%>%
               group_by(interval)%>%
               summarise(mean_per_interval = mean(steps))

plot(x=steps_per_interval$interval, y=steps_per_interval$mean_per_interval, type="l", xlab = "5 Minute intervals throughout the day", ylab = "Mean Steps Taken per 5 Minute Interval") 
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
``` {r, echo=TRUE}
steps_per_interval[steps_per_interval$mean_per_interval==max(steps_per_interval$mean_per_interval),]

```


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo = TRUE}
sum(is.na(activity$steps))

#or as a percentage of total observations
paste(round(sum(is.na(activity$steps))/length(activity$steps)*100,2),"%", sep="")

```
2. Devise a strategy for filling in all of the missing values in the dataset. 

We will be using the median values per 5 minute interval & which day of the week to fill in mising values

``` {r, echo = TRUE}

#add days of week to our data

activity<-activity%>%mutate(day_of_week=wday(date, label=TRUE))

#replace 
activity2<-activity %>%
     group_by(interval, day_of_week) %>% 
     mutate_each(funs(replace(., which(is.na(.)),
                                mean(., na.rm=TRUE))))

```

### Based on the new data set, repeat our analysis above

```{r, echo = TRUE}
steps_per_day2<-activity2%>%
               group_by(date)%>%
               summarise(total_no_steps_per_day = sum(steps))

steps_per_day2

#histogram of the new "complete" dataset
qplot(steps_per_day2$total_no_steps_per_day, geom="histogram") + labs(x = "Total Number of Steps per Day")


#density
ggplot(steps_per_day2, aes(x=total_no_steps_per_day)) + geom_density() + labs(x = "Total Number of Steps Per Day")

#mean steps taken each day
mean(steps_per_day2$total_no_steps_per_day)

#median steps taken each day
median(steps_per_day2$total_no_steps_per_day)

```
Comparing the original dataset vs the dataset with missing values imputed, we can see minor differences in the mean (original: `r round(mean(steps_per_day$total_no_steps_per_day),2)`; imputed: `r round(mean(steps_per_day2$total_no_steps_per_day),2)`) and median (original: `r round(median(steps_per_day$total_no_steps_per_day),2)`; imputed: `r round(median(steps_per_day2$total_no_steps_per_day),2)`)

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}

#bucket days of the week into two factors; weekend & weekday
activity2$weekend <- chron::is.weekend(activity2$date)
activity2$weekend[activity2$weekend==TRUE]<-"weekend"
activity2$weekend[activity2$weekend==FALSE]<-"weekday"

steps_per_interval_weekends<-activity2%>%
               group_by(interval, weekend)%>%
               summarise(mean_per_interval = mean(steps))

ggplot(steps_per_interval_weekends, aes(x=interval, y=mean_per_interval)) + geom_line() + facet_wrap(~weekend) + labs(x = "Interval periods (5 minute blocks)", y = "Average Number of steps per interval")

```
